The Power of Two Random Choices

### 2014-03-24 presentation/presentation.tex
some minor fixes
more slides

\AtBeginSubsubsection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[subsubsectionstyle=show/shaded,currentsection,currentsubsection]
  \end{frame}
}

@@

\AtBeginSection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[subsubsectionstyle=show/shaded,currentsection,currentsubsection]
  \end{frame}
}

%---------------------------------------------------------------------
% Create Color definition From Template
%---------------------------------------------------------------------
% #1 template name, 
% #2 foreground color name
% #3 background color name
\newcommand{\ccft}[3]{
\usebeamercolor{#1}
\definecolor{#2}{named}{fg}
\definecolor{#3}{named}{bg}
}
%get some colors from the scheme
\ccft{block body}{myblock body fg}{myblock body bg}
\ccft{block title}{myblock title fg}{myblock title bg}

%---------------------------------------------------------------------
%parameters for balls into bins graphs
%---------------------------------------------------------------------
%scale factor
\newcommand\scalefac{0.55}
%diameter of balls
\newcommand\ballsize{5mm}
%number of bins
\newcommand\nrbins{6}
%space between ball and bin
\newcommand\padding{0.1*\ballsize}
%height of a bin
\newcommand\binheight{5*\balldiameter}
%width of a bin
\newcommand\binwidth{\balldiameter}
%gap from bin to bin
\newcommand\bingap{1.6*\balldiameter}
%color of balls
\tikzstyle{ballstyle} = [ball color=black!30!red]

@@

\section{Balls into bins}
\begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents
\end{frame}

\begin{frame}
\frametitle{The Problem}
\begin{block}{Balls-into-bins games}
Suppose we have $n$ initially empty bins and need to distribute $m \geq n$ balls among them, as evenly as possible. We are interested in the maximum load $\load$, i.e, how many balls are in the fullest bin?
\end{block}
\pause
Applications:
\begin{itemize}
	\item Hashing
	\item Online load balancing
	\item Emulating PRAMs on DMMs
	\item Low congestion circuit routing
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Single choice}
\framesubtitle{A basic allocation scheme}
\begin{block}{Single choice strategy}
For each ball, choose one of the $n$ bins uniformly and independently at random.
\end{block}
\bigskip
\begin{center}
\begin{tikzpicture}[scale=\scalefac]
	\bab{2 1 5 2 3 0}
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Randomized allocation process}

Since we deal with a randomized process, we cannot give upper bounds for the maximum load $\load$ with absolute certainty.

\bigskip

\begin{block}{Definition}
We say that an event $\mathcal E_n$ occurs \emph{with high probability} (w.h.p.) if $\Pr\left[\mathcal E_n \right] \geq 1 - n^{-\alpha}$ for some $\alpha > 0$.
\end{block}
\end{frame}

\begin{frame}
\frametitle{Single choice}
\framesubtitle{Upper bound for the maximum load}
\begin{claim}
When using the single choice strategy to distribute $m$ balls to $n$ bins, the maximum load will be, w.h.p.,
\begin{align}
\load = 
	\begin{cases}
    \frac{\ln n}{\ln\ln n} + O(1)              & m = n \\
    \frac{m}{n} + \Theta\left(\sqrt{\frac{m\ln n}{n}} \right)              & m > n
    \end{cases}
\end{align}
\end{claim}
\end{frame}

\begin{comment}
\begin{frame}
\frametitle{Applications}
\begin{itemize}
	\item Hashing
	\item Online load balancing
	\item Emulating PRAMs on DMMs
	\item Low congestion circuit routing
\end{itemize}
\end{frame}
\end{comment}

%Table of contents at the start of each section
\AtBeginSection[]
{
  \begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents[currentsection]
  \end{frame}
}

\section{Allocation shemes}
\begin{frame}
\frametitle{BiB Model}
\end{frame}

\begin{frame}
\frametitle{Multiple choice strategies}
\end{frame}

\begin{frame}
\frametitle{Greedy}
\end{frame}

\begin{frame}
\frametitle{Always-go-left}
\end{frame}

@@

@@

\begin{comment}
\begin{frame}[label=references]
  \frametitle{References}
  %\nocite{*}
  %\setbeamertemplate{bibliography item}[book]
  \bibliographystyle{acm}
  \bibliography{../sources}
\end{frame}
\note{}
\end{comment}

\begin{frame}{Introduction}
\end{frame}
\note{}

@@

@@

\begin{frame}{Title}
\end{frame}
\note{}

>>>>

\begin{comment}
\AtBeginSubsubsection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[subsubsectionstyle=show/shaded,currentsection,currentsubsection]
  \end{frame}
}

@@

\AtBeginSection[]
{
  \begin{frame}<beamer>
    \frametitle{Outline}
    \tableofcontents[subsubsectionstyle=show/shaded,currentsection,currentsubsection]
  \end{frame}
}
\end{comment}

%---------------------------------------------------------------------
% Create Color definition From Template
%---------------------------------------------------------------------
% #1 template name, 
% #2 foreground color name
% #3 background color name
\newcommand{\ccft}[3]{
\usebeamercolor{#1}
\definecolor{#2}{named}{fg}
\definecolor{#3}{named}{bg}
}
%get some colors from the scheme
\ccft{block body example}{myblock body fg}{myblock body bg} 
%---------------------------------------------------------------------
%parameters for balls into bins graphs
%---------------------------------------------------------------------
%scale factor
\newcommand\scalefac{0.55}
%diameter of balls
\newcommand\ballsize{5mm}
%number of bins
\newcommand\nrbins{6}
%space between ball and bin
\newcommand\padding{0.1*\ballsize}
%height of a bin
\newcommand\binheight{5*\balldiameter}
%width of a bin
\newcommand\binwidth{\balldiameter}
%gap from bin to bin
\newcommand\bingap{1.6*\balldiameter}
%color of balls
\tikzstyle{ballstyle} = [ball color=black!30!red]

@@

\begin{comment}
\begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents
\end{frame}
\end{comment}

\section{Balls into bins}
\begin{frame}
\frametitle{The Problem}
\begin{exampleblock}{Balls-into-bins games}
Suppose we have $n$ initially empty bins and need to distribute $m \geq n$ balls among them, \alert{as evenly as possible}. 
\end{exampleblock}

\begin{exampleblock}{Definition}
We are interested in the \alert{maximum load $\load$}, i.e, how many balls are in the fullest bin?
\end{exampleblock}

\begin{exampleblock}{Definition}
Equivalently, we can look at the \alert{additive gap} $\delta$:
\begin{align*}
\delta = \load - \frac{m}{n}
\end{align*}
\end{exampleblock}

\end{frame}

\begin{frame}
\frametitle{Applications}
Applications:
\begin{itemize}
	\item Hashing
	\item Online load balancing
	\item Emulating PRAMs on DMMs
	\item Low congestion circuit routing
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Single choice}
\framesubtitle{A basic allocation scheme}
\begin{exampleblock}{Single choice strategy}
For each ball, choose one of the $n$ bins uniformly and independently at random.
\end{exampleblock}
\bigskip
\begin{center}
\begin{tikzpicture}[scale=\scalefac]
	\bab{2 1 5 2 3 0}
\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Some definitions}

Since we deal with a \alert{randomized process}, we cannot give upper bounds for the maximum load $\load$ with absolute certainty.

\begin{exampleblock}{Definition}
We say that an event $\mathcal E_n$ occurs \alert{with high probability} (w.h.p.) if $\Pr\left[\mathcal E_n \right] \geq 1 - n^{-\alpha}$ for some $\alpha > 0$.
\end{exampleblock}

\bigskip
\pause

%\begin{exampleblock}{Definition}
We distinguish:
\begin{itemize}
\item The \alert{lightly loaded} case: $m = n$
\item The \alert{heavily loaded} case:	 $m \in \omega(n\ln n)$
\end{itemize}
%\end{exampleblock}

\end{frame}

\begin{frame}
\frametitle{Single choice}
\framesubtitle{Upper bound for the maximum load}
\begin{theorem}[Raab and Steger, 1998]
When using the single choice strategy to distribute $m$ balls to $n$ bins, the maximum load will be, w.h.p.,
\begin{align*}
\load = 
	\begin{cases}
    \frac{\ln n}{\ln\ln n} + O(1)              & m = n \\
    \frac{m}{n} + \Theta\left(\sqrt{\frac{m\ln n}{n}} \right)              & m \in \omega(n \ln n)
    \end{cases}
\end{align*}
\end{theorem}
\end{frame}

%Table of contents at the start of each section
\begin{comment}
\AtBeginSection[]
{
  \begin{frame}
    \frametitle{Table of Contents}
    \tableofcontents[currentsection]
  \end{frame}
}
\end{comment}

\section{Allocation schemes}
\begin{frame}
\frametitle{Multiple choice strategies}

\begin{exampleblock}{Idea}
Put a ball into the \alert{least full} of $d\geq2$ randomly chosen bins.
\end{exampleblock}

\pause
\bigskip
We can distinguish placement algorithms by the way the $d$ \alert{bins are sampled}:
\begin{itemize}
\item Uniformly and independently
\item Non-unfiromly and independently
\item Non-uniformly and dependently
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Greedy}
\begin{exampleblock}{The Greedy algorithm}
\begin{enumerate}
\item For each ball, choose $d$ bins independently and uniformly at random. 
\item Put the ball into the least full bin.
\item In case a tie occurs, choose one of the $d$ bins arbitrarily.
\end{enumerate}
\end{exampleblock}
\todo{visualize}
\end{frame}

\begin{frame}
\frametitle{Greedy}
\begin{theorem}[Azar, Broder, Karlin and Upfal, 1999]
When using the \alert{Greedy} algorithm to distribute $m$ balls to $n$ bins, the maximum load will be (w.h.p.~and for all $m$): 
\begin{align*}
\load = \frac{m}{n} + \frac{\ln \ln n}{\ln d} + O(1) \quad .
\end{align*}
\end{theorem}
\pause
\medskip
Compared to the \alert{single choice} strategy, with 
\begin{align*}
\load = 
	\begin{cases}
    \frac{\ln n}{\ln\ln n} + O(1)              & m = n \\
    \frac{m}{n} + \Theta\left(\sqrt{\frac{m\ln n}{n}} \right)              & m \in \omega(n \ln n)
    \end{cases} \quad ,
\end{align*}
\alert{Greedy} with $d=2$:
\begin{itemize}
\item achieves an \alert{exponential decrease} of the additive gap $\delta$ and
\item $\delta$ does \alert{not depend on} the number of balls \alert{$m$} anymore.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Always-go-left}
\begin{exampleblock}{The Always-go-left algorithm}
\begin{enumerate}
\item Partition the $n$ bins into $d$ groups.
\item Assume a fixed ordering of the bins $b_i$, $1\leq i \leq n$.
\item Choose one bin from each group uniformly and independently.
\item Put the ball into the least full bin.
\item In case a tie occurs, put the ball into the leftmost bin (i.e., the bin with smallest index) among those with same minimal load.
\end{enumerate}
\end{exampleblock}
\todo{visualize}
\end{frame}

\begin{frame}
\frametitle{Always-go-left}
\begin{theorem}[V\"ocking, 2003]
When using the \alert{Always-go-left} algorithm to distribute $m$ balls to $n$ bins, the maximum load will be  (w.h.p.~and for all $m$): 
\begin{align*}
\load = \frac{m}{n} + \frac{\ln \ln n}{d \ln \Phi_d} + O(1) \quad .
\end{align*}
\end{theorem}
\todo{$\Phi_d$ is undefined. I think a full definition only makes sense if we present the proof for Always-go-left.}

\bigskip
The combination of \alert{asymmetric tie breaking} and \alert{non-uniform} (and independent) sampling of the $d$ bins is crucial to achieve this upper bound.
\end{frame}

\begin{frame}
\frametitle{BiB Model}
We will assume a sequential, infinite, dynamic process.

\begin{description}
\item[\alert{Sequential:}] The balls arrive one after another.
\item[\alert{Infinite:}] Ball placements and deletions are given for an infinite time span.
\item[\alert{Dynamic:}] At each time step $t$, one ball is removed from the bins.
\begin{itemize}
\item An oblivious adversary provides a sequence $\sigma = \sigma_1 \sigma_2 \dots$, where $\sigma_t = (d_t, a_t)$.
\item $d_t$/$a_t$: ball to delete/add at time step t. \\
\item $\sigma$ is fixed in advance.
\end{itemize}
\end{description}
\end{frame}

@@

\begin{frame}
\frametitle{Conclusion}
\end{frame}

@@

@@

\begin{frame}[allowframebreaks, label=references]
\frametitle<presentation>{References}    
\bibliographystyle{acm}
\nocite{*}
\setbeamertemplate{bibliography item}[text]
\bibliography{../sources}
\end{frame}

@@
